---
title: "A Look at the DAO Case Funnel"
author: "DAO DATA Lab"
date: 11/4/2019
output:
  html_notebook
---

The DATA Lab's first data story based off of the [DAO Data Dashboard](https://data.philadao.com).  The goal is to look at the funnel from Incidents to Arrests to Charges.  Usually this is a huge funnel - there are many more incidents than Arrests and a small percentage more arrests than cases Charged.  In some instances, though, like drug offenses, that funnel can be inverted.  This phenomenon is discussed more in depth in our Medium story, which can also be found on the dashboard.  

```{r include = FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(stringi)

# download the arrests CSV
arrests <- read.csv("https://data.philadao.com/data/arrest_data_daily_citywide.csv", stringsAsFactors = FALSE) %>%
  # make the data "tidy"/tall
  gather(Offense, value, -date_value) %>%
  # replace "." with spaces in the offense names
  mutate(Offense = gsub("\\.", " ", Offense),
         # turn the date into a date
         date_value = as.Date(date_value),
         # create a new column that represents the month that an event took place
         month = floor_date(date_value, "month"))

# download the charges CSV
charges <- read.csv("https://data.philadao.com/data/charges_data_daily_citywide.csv", stringsAsFactors = FALSE) %>%
  # make the data "tidy"/tall  
  gather(Offense, value, -date_value) %>%
  # replace "." with spaces in the offense names
  mutate(Offense = gsub("\\.", " ", Offense),
         # turn the date into a date objedct
         date_value = as.Date(date_value),
         # create a new column that represents the month that an event took place
         month = floor_date(date_value, "month"))
```

```{r include = FALSE}
# this function is taken from our cache creation code, but is put in here for public consumption.
# it downloads all incident data from open data philly based on the start and end dates supplied
getIncidentData <- function(s_date, e_date) {
  url_base <- "https://phl.carto.com/api/v2/sql?q="
  url_select <- "SELECT+objectid,+dc_dist,+psa,+dispatch_date_time,+dispatch_time,+dc_key,+location_block,+ucr_general,text_general_code,+hour_,+ST_Y(the_geom)+AS+lat,+ST_X(the_geom)+AS+lng+FROM+incidents_part1_part2"
  url_where <- paste0("+WHERE+dispatch_date_time+BETWEEN+%27", s_date,"%27+AND+%27", as.character(e_date), "%27")
  url_filename <- "&filename=incidents_part1_part2"
  url_format <- "&format=csv"
  url_skip <- "&skipfields=cartodb_id"
  
  # read csv from url constructed above
  incidents <- read.csv(paste0(url_base, url_select, url_where, url_filename, url_format, url_skip), stringsAsFactors = FALSE)
  
  # create and format a number of fields: 
  # create common dc_number field for joining to PARS and DAOCMS data, format date/datetime fields
  incidents <- incidents %>%
    mutate(dc_number = substring(dc_key, 3),
           dispatch_date_time = as.POSIXct(dispatch_date_time),
           dispatch_date = as.Date(dispatch_date_time),
           dc_dist = stri_pad(as.character(dc_dist), width = 2, pad = "0", side = "left"),
           month = floor_date(dispatch_date, "month")) %>%
    rename(dc_district = dc_dist)
  
  # reorder columns logically
  incidents <- incidents[c("objectid", "dc_number", "dc_key", "dispatch_date_time", "dispatch_date",
                           "month", "dispatch_time", "ucr_general", "text_general_code", "dc_district",
                           "location_block", "psa", "hour_", "lat", "lng")]
  
  return(incidents)

}

# used to shade the graphs
krasnerInOffice <- tibble(
  start = c(ymd("2018-01-02")),
  end = c(today())
)


# graphs just charges; not used in this bit of code anymore, so not commented well
graph_charges <- function(df, prefix, plot_title, plot_vline_date = NA, plot_vline_label = NA) {
  cases <- df %>% 
    select(date_value, starts_with(prefix)) %>%
    gather(Offense, value, -date_value) %>%
    mutate(Offense = gsub("\\.", " ", Offense),
           date_value = as.Date(date_value),
           month = floor_date(date_value, "month")) %>%
    filter(month < max(month)) %>%
    group_by(month, Offense) %>%
    summarize(
      `Total Cases Charged` = sum(value),
    )

  plot <- cases %>% 
    ggplot(aes(x = month, y = `Total Cases Charged`, color = Offense))
  
  if(!is.na(plot_vline_date)) {
    plot <- plot + geom_vline(xintercept = plot_vline_date, color = 'grey25', size = 1) +
      geom_text(aes(x = plot_vline_date, label = plot_vline_label, y = 0), colour='grey25',
              angle=90, hjust = 0, vjust = -.4, size=3.5)
  }
  
  plot <- plot + 
    geom_rect(data = krasnerInOffice, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf),
      fill = "pink", alpha = 0.5, inherit.aes = FALSE
    ) +
    scale_x_date(expand = expand_scale(mult = c(0.04, 0))) + 
    geom_point() +
    geom_smooth(method = "loess", span=.5, se=FALSE, size = 1) + 
    scale_y_continuous("Cases Charged",
                       limits = c(0, NA)) + 
    labs(title = plot_title,
         y = "Count Per Month") + 
    theme_minimal() + 
    theme(axis.title.x=element_blank())

  
  # remove legend if there is only one grouping; move it to the bottom otherwise
  if(n_distinct(cases$Offense) > 1) {
    plot <- plot + theme(legend.position="bottom")
  } else {
    plot <- plot + theme(legend.position="none")
  } 
  
  return(plot)
}

# graphs the funnel of incidents, arrests, and charges; takes a plot title
graph_incidents_arrests_charges <- function(df_incidents, df_arrests, df_charges, plot_title) {
  monthly_df <- df_incidents %>%
    # start by greating a monthly count of each data set and left_joining them together by month
    count(month, name = "Incidents") %>% 
    left_join(df_arrests %>% group_by(month) %>% summarize(Arrests = sum(value)),
              by = "month") %>%
    left_join(df_charges %>% group_by(month) %>% summarize(Charges = sum(value)),
              by = "month") %>%
    # make the data tidy
    gather(type, value, -month) %>%
    # get rid of the most recent month since it may not be a complete month
    filter(month < max(month))
  
  # create the plot
  plot <- monthly_df %>%
    ggplot(aes(x = month, y = value, color = type)) +
    # put in a shading for the current DA
    geom_rect(data = krasnerInOffice, aes(xmin = start, xmax = end, ymin = -Inf, ymax = +Inf),
      fill = "pink", alpha = 0.5, inherit.aes = FALSE
    ) +
    # scaling so that our dots don't touch the edge of the graph
    scale_x_date(expand = expand_scale(mult = c(0.04, 0))) + 
    geom_point() +
    # add a loess curve
    geom_smooth(method = "loess", span=.5, se=FALSE, size = 1) + 
    #labels
    labs(title = plot_title,
         x = "",
         y = "Count Per Month",
         color = "") + 
    # remove most elements from the plot for a clean finish
    theme_minimal()  

  return(plot)
}
```

```{r, include=FALSE}
incidents <- getIncidentData("2014-01-01", today())
```



**Graph of the overall number of Incidents, Arrests, and Charges per month**
```{r, echo = FALSE}
graph_incidents_arrests_charges(incidents, arrests, charges, "The Funnel from Incidents to Charges")
ggsave("funnel_incidents_charges.png", width=6, height=4)
```

**Graph of the number of Incidents, Arrests, and Charges per month for Drug Offenses**
```{r, echo = FALSE}
graph_incidents_arrests_charges(incidents %>% filter(ucr_general==1800), arrests %>% filter(startsWith(Offense, "Drug")), charges %>% filter(startsWith(Offense, "Drug")), "The Funnel from Incidents to Charges:\nDrug Cases")

ggsave("funnel_incidents_charges_drugs.png", width=6, height=4)
```

**Graph of the number of Incidents, Arrests, and Charges per month for Robbery With a Firearm Offenses**
```{r, echo = FALSE}
graph_incidents_arrests_charges(incidents %>% filter(text_general_code == "Robbery Firearm"), arrests %>% filter(Offense == "Robbery Gun"), charges %>% filter(Offense == "Robbery Gun"), "The Funnel from Incidents to Charges: \nRobbery w/Gun Cases")

ggsave("funnel_incidents_charges_robbery_gun.png", width=6, height=4)
```
