---
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(conflicted)

library(lubridate)

filter <- dplyr::filter

knitr::opts_chunk$set(echo = TRUE)
```

# Estimating the Number of Intravenous Drug Users (IDUs) in Phildelphia

## Lower Bound IDU Estimate

For a lower-bound estimate of the number of intravenous / injectable drug users (IDUs) in the city, I'll simply add the number of individuals arrested for IDU + the number of overdoses + the number of people in treatment. For the moment, intravenous drug users are defined as users of heroin, fentanyl or fentanyl analogues.

### Read In / Wrangle Data

#### Arrests

From Philly Open Data <https://github.com/phillydao/phillydao-public-data/tree/master/docs>

Read in the data

```{r}
### read in the table from Philly Open Data for daily arrests
vroom::vroom('phillydao-public_data/docs/data/arrest_data_daily_by_zip.csv') -> daily_arrests_zip

### read in the table for drug seizures; going to assume people are arrested
### for using drugs in the same proportion as those drugs are seized
vroom::vroom('philly_doh_tableau_data/drug_seizures.tsv') %>%
  mutate(quarter = case_when(quarter == 'Q1' ~ '01-01',
                             quarter == 'Q2' ~ '04-01',
                             quarter == 'Q3' ~ '07-01',
                             quarter == 'Q4' ~ '10-01')) %>%
  unite(quarter, c('year', 'quarter'), sep = '-') %>%
  mutate(quarter = as.Date(quarter)) -> drug_seizures
```

Summarize arrests to the year and quarter (since that's how fine-grained some of the other data is) and estimate IUDs by proportion of drugs seized.

```{r}
daily_arrests_zip %>%
  select(date_value, `Drug Possession`) %>%
  mutate(year = year(date_value),
         quarter = round_date(date_value, unit = 'quarter')) %>%
  group_by(year, quarter) %>%
  summarize(arrests = sum(`Drug Possession`)) %>%
  left_join(drug_seizures) %>%
  mutate(iud_perc = fentanyl + fentanyl_analogue + heroin) %>%
  select(-(cocaine:other)) %>%
  mutate(arrests_adj = round(arrests * (iud_perc / 100))) -> arrests
```

#### Overdoses

From Philadelphia Department of Health's public opiod tableau page <https://public.tableau.com/profile/pdph#!/>

```{r}
vroom::vroom('philly_doh_tableau_data/overdose_deaths.tsv') -> overdose_deaths
vroom::vroom('philly_doh_tableau_data/hosp_nonfatal_overdoses.tsv') -> nonfatal_over

### This is a combination of overdose deaths and nonfatal overdose, but far 
### fewer years of datacompared to to the tables above so decided not to use
#vroom::vroom('../../../data/data/dph_od_incident_counts.csv') -> overdose_incidents

### Decided not to use this table because it's voluntarily entered first 
### responder information and is probably counted in the other dataset
#vroom::vroom('../../../data/psp_overdose_events.csv') -> overdoses
```

Combine the data

```{r}
overdose_deaths %>%
  filter(year != 2019) %>%
  group_by(year) %>%
  summarize(deaths = sum(deaths)) %>%
  ungroup() %>%
# add number from city report for 2019 here
# https://www.phila.gov/media/20200226121229/Substance-Abuse-Data-Report-02.26.20.pdf
  rbind(tibble(year = 2019, deaths = 950)) %>%
  left_join(nonfatal_over) %>%
  mutate(overdoses = deaths + cases) %>%
  select(year, overdoses) -> overdoses
```

#### Individuals in Medication Assisted Treatment

From Philadelphia Department of Health's public opiod tableau page <https://public.tableau.com/profile/pdph#!/>

```{r}
vroom::vroom('philly_doh_tableau_data/medicaid_beneficiaries_in_medication_assisted_treatment.tsv') -> mat_medicaid
vroom::vroom('philly_doh_tableau_data/prisoners_in_medication_assisted_treatment.tsv') -> mat_prison
```

Combine the data

```{r}
mat_medicaid %>%
  group_by(year) %>%
  summarize(patients = sum(individuals)) %>%
  ungroup() %>%
  left_join(mat_prison) %>%
  replace(is.na(.), 0) %>%
  mutate(mat_patients = patients + individuals) %>%
  select(year, mat_patients) -> mat
```

#### Combine 3 Types of  Data

```{r}
arrests %>%
  group_by(year) %>%
  summarise(arrests = sum(arrests),
            iud_perc = mean(iud_perc),
            arrests_adj = round(arrests * (iud_perc / 100))) %>%
  left_join(overdoses) %>%
  left_join(mat) %>%
  mutate(lb = arrests + overdoses + mat_patients) %>% #lm(lb ~ year, data = .) %>% summary()
  rbind(tibble(year = 2021:2025,
               arrests = NA,
               iud_perc = NA, 
               arrests_adj = NA,
               overdoses = NA, 
               mat_patients = NA,
               lb = NA)) %>%
  mutate(est_lb = round((year * 1056.6) - 2115104.8)) -> data
```

<br>

---

<br>

### Visualize

Data used to construct the estimate.

```{r}
data %>%
  select(-iud_perc, -arrests_adj, -lb, -est_lb) %>%
  pivot_longer(arrests:mat_patients, 
               names_to = 'data_source', 
               values_to = 'count') %>%
  filter(year != 2020) %>%
  na.omit() -> raw_data_skinny
  
ggplot(raw_data_skinny, aes(x = year, y = count, color = data_source)) +
  geom_line() +
  geom_point() +
  scale_color_manual(name = 'Data Source',
                     values = c('dodgerblue3', 'orange3', 'forestgreen')) +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) 
ggsave('addition_method_opioid_users.png')
```

Plot estimated number of opiod users.

```{r, fig.width = 5, fig.height = 4}
data %>%
  select(year, lb, est_lb) %>%
  mutate(est_lb = ifelse(is.na(lb) == F, lb, est_lb)) %>%
  pivot_longer(lb:est_lb, 
               names_to = 'estimate_source', values_to = 'estimate') %>%

ggplot(aes(x = year, y = estimate, color = estimate_source)) +
  geom_line(linetype = rep(c('dashed', 'solid'), each = 12)) +
  scale_color_manual(values = c('black', 'black')) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20)  +
  theme(legend.position = 'none')
ggsave('addition_opiod_user_estimate.png')
```

#### Adjusted Estimate

- ~32% of people in treatment say they have injected drugs
- ~46% - 60% based on HCV data

```{r}
data %>%
  select(year, lb, est_lb) %>%
  mutate(upper_hcv_adjust = round(ifelse(is.na(lb) == F, 
                                         lb * 0.6, est_lb * 0.6)),
         middle_hcv_adjust = round(ifelse(is.na(lb) == F, 
                                         lb * 0.46, est_lb * 0.46)),
         low_treat_adjust  = round(ifelse(is.na(lb) == F, 
                                         lb * 0.32, est_lb * 0.32)),
         ymin = low_treat_adjust,
         ymax = upper_hcv_adjust) %>%
  pivot_longer(names_to = 'adjustment', values_to = 'estimate', 
               upper_hcv_adjust:low_treat_adjust) %>%

ggplot(aes(x = year, y = estimate, color = adjustment)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.15, color = 'gray60') +
  geom_line(aes(linetype = adjustment)) +
  geom_point() +
  scale_color_manual(values = rep('black', 3)) +
  scale_linetype_manual(values = c('dashed', 'solid', 'dashed')) +
  labs(x = 'Year', y = 'Estimated Number of\nIV Drug Users') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
#ggsave('estimated_iud_users.png')
```


## Multiplier Method using Cinncinati Paper

<https://jheor.org/article/9729-estimating-the-prevalence-of-opioid-use-disorder-in-the-cincinnati-region-using-probabilistic-multiplier-methods-and-model-averaging>

### Calculate Multipliers

#### Fatal Overdose

Multiplier: 7.182 %
Std. Error: 0.533 %

```{r}
overdose_deaths %>%
  mutate(se_lower = round((deaths * 100) / (7.182 - 0.533)),
         multiplier = round((deaths * 100) / 7.182),
         se_upper = round((deaths * 100) / (7.182 + 0.533))) %>%
  group_by(year) %>%
  summarize(sum_multiplier = sum(multiplier),
            sum_se_low = sum(se_lower),
            sum_se_high = sum(se_upper)) %>%
  ungroup() %>% #lm(sum_multiplier ~ year, data = .) %>% broom::tidy()
  filter(year != '2019') %>%
  rbind(tibble(year = 2019:2025,
               sum_multiplier = NA,
               sum_se_low = NA,
               sum_se_high = NA)) %>%
  mutate(projection = ifelse(year < 2019, sum_multiplier,
                             round(((year * 654.2) + -1310155.4))),
         proj_se_low = ifelse(year < 2019, sum_se_low,
                              round((projection - (1.96 * 395.8294)))),
         proj_se_high = ifelse(year < 2019, sum_se_high,
                              round((projection + (1.96 * 395.8294))))) %>%
  unite(multiplier, sum_multiplier:sum_se_high, sep = '_') %>%
  unite(projection, projection:proj_se_high) %>%
  pivot_longer(multiplier:projection,
               names_to = 'estimate_source', values_to = 'estimate') %>%
  separate(estimate, into = c('estimate', 'low', 'high'), 
           sep = '_', convert = T) %>%
  na.omit() %>%
  mutate(data_source = 'fatal_overdoses') -> fatal_overdose_multiplier
```

#### Treatment Admission

Multiplier: 21.376 %
Std. Error: 2.142 %

```{r}
mat %>%
  mutate(se_lower = round((mat_patients * 100) / (21.376 - 2.142)),
         multiplier = round((mat_patients * 100) / 21.376),
         se_upper = round((mat_patients * 100) / (21.376 + 2.142))) %>%
  group_by(year) %>%
  summarize(sum_multiplier = sum(multiplier),
            sum_se_low = sum(se_lower),
            sum_se_high = sum(se_upper)) %>%
  ungroup() %>% #lm(sum_multiplier ~ year, data = .) %>% broom::tidy()
  filter(year != '2019') %>%
  rbind(tibble(year = 2019:2025,
               sum_multiplier = NA,
               sum_se_low = NA,
               sum_se_high = NA)) %>%
  mutate(projection = ifelse(year < 2019, sum_multiplier,
                             round(((year * 3494.317) + -7006229.989))),
         proj_se_low = ifelse(year < 2019, sum_se_low,
                              round((projection - (1.96 * 447.1714)))),
         proj_se_high = ifelse(year < 2019, sum_se_high,
                              round((projection + (1.96 * 447.1714))))) %>%
  unite(multiplier, sum_multiplier:sum_se_high, sep = '_') %>%
  unite(projection, projection:proj_se_high) %>%
  pivot_longer(multiplier:projection,
               names_to = 'estimate_source', values_to = 'estimate') %>%
  separate(estimate, into = c('estimate', 'low', 'high'), 
           sep = '_', convert = T) %>%
  na.omit() %>%
  mutate(data_source = 'MAT') -> mat_multiplier
```

#### Non-Fatal ED Overdose

Multiplier: 40.890 %
Std. Error: 5.111 %

```{r}
overdoses %>%
  mutate(se_lower = round((overdoses * 100) / (21.376 - 2.142)),
         multiplier = round((overdoses * 100) / 21.376),
         se_upper = round((overdoses * 100) / (21.376 + 2.142))) %>%
  group_by(year) %>%
  summarize(sum_multiplier = sum(multiplier),
            sum_se_low = sum(se_lower),
            sum_se_high = sum(se_upper)) %>%
  ungroup() %>% #lm(sum_multiplier ~ year, data = .) %>% broom::tidy()
  filter(year != '2019') %>%
  rbind(tibble(year = 2019:2025,
               sum_multiplier = NA,
               sum_se_low = NA,
               sum_se_high = NA)) %>%
  mutate(projection = ifelse(year < 2019, sum_multiplier,
                             round(((year * 554.6833) + -1111428.5667))),
         proj_se_low = ifelse(year < 2019, sum_se_low,
                              round((projection - (1.96 * 83.90343)))),
         proj_se_high = ifelse(year < 2019, sum_se_high,
                              round((projection + (1.96 * 83.90343))))) %>%
  unite(multiplier, sum_multiplier:sum_se_high, sep = '_') %>%
  unite(projection, projection:proj_se_high) %>%
  pivot_longer(multiplier:projection,
               names_to = 'estimate_source', values_to = 'estimate') %>%
  separate(estimate, into = c('estimate', 'low', 'high'), 
           sep = '_', convert = T) %>%
  na.omit() %>%
  mutate(data_source = 'nonfatal_overdoses') -> nonfatal_overdose_multiplier
```

#### Combine Data

```{r}
rbind(fatal_overdose_multiplier, 
      nonfatal_overdose_multiplier, 
      mat_multiplier) %>%
  unite(group, c('estimate_source', 'data_source'), 
        sep = '_', remove = F)-> multipliers
```

#### Plot

```{r, fig.width = 5, fig.height = 4}
### all three multipliers
multipliers %>%

ggplot(aes(x = year, y = estimate, color = group, group = group)) +
  geom_ribbon(aes(ymin = low, ymax = high, fill = group), 
              alpha = 0.15, color = 'white') +
  scale_fill_manual(name = '', values = rep(c('firebrick3', 'orange3', 'pink3'), 2)) +
  ggrepel::geom_label_repel(data = filter(multipliers, year == 2025),
                           aes(label = data_source, 
                               x = 2026, 
                               # fatal overdoses, nonfatal overdoses, MAT
                               y = c(18000, 5000, 60000))) +
  geom_line(aes(linetype = group)) +
  scale_linetype_manual(name = '', values = rep(c('solid', 'dashed'), each = 3)) +
  scale_color_manual(name = '', 
                     values = rep(c('firebrick3', 'orange3', 'pink3'), 2)) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
#ggsave('opioid_users_all_multiplier_estimates.png')

### without MAT
multipliers %>%
  filter(data_source != 'MAT') %>%

ggplot(aes(x = year, y = estimate, color = group, group = group)) +
  geom_ribbon(aes(ymin = low, ymax = high, fill = group), 
              alpha = 0.15, color = 'white') +
  scale_fill_manual(name = '', values = rep(c('firebrick3', 'pink3'), 2)) +
  ggrepel::geom_label_repel(data = filter(multipliers, year == 2025,
                                          data_source != 'MAT'),
                           aes(label = data_source, 
                               x = 2026, 
                               # fatal overdoses, nonfatal overdoses, MAT
                               y = c(15500, 7000))) +
  geom_line(aes(linetype = group)) +
  scale_linetype_manual(name = '', 
                        values = rep(c('solid', 'dashed'), each = 2)) +
  scale_color_manual(name = '', 
                     values = rep(c('firebrick3', 'pink3'), 2)) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
  theme(legend.position = 'none')
#ggsave('opioid_users_multiplier_estimates_no_MAT.png')
```

### Adjusted Multiplier Estimates for IDUs

#### Adjusted Estimate

- ~32% of people in treatment say they have injected drugs
- ~46% - 60% based on HCV data

```{r}
multipliers %>%
  mutate(low_adj_est = estimate * 0.32,
         low_adj_low = low * 0.32,
         low_adj_high = high * 0.32,
         mid_adj_est = estimate * 0.46,
         mid_adj_low = low * 0.46,
         mid_adj_high = high * 0.46,
         high_adj_est = estimate * 0.6,
         high_adj_low = low * 0.6,
         high_adj_high = high * 0.6) %>%
  group_by(year, group, estimate_source, data_source) %>%
  summarize(adj_estimate = (low_adj_est + mid_adj_est + high_adj_est) / 3,
            adj_low = min(low_adj_low, mid_adj_low, high_adj_low),
            adj_high = max(low_adj_high, mid_adj_high, high_adj_high)) %>%
  ungroup() -> multipliers_adjusted
```

#### Plot Adjusted Estimate

```{r, fig.width = 5, fig.height = 4}
### all three multipliers
multipliers_adjusted %>%

ggplot(aes(x = year, y = adj_estimate, color = group, group = group)) +
  geom_ribbon(aes(ymin = adj_low, ymax = adj_high, fill = group), 
              alpha = 0.15, color = 'white') +
  scale_fill_manual(name = '', values = rep(c('firebrick3', 'orange3', 'pink3'), 2)) +
  ggrepel::geom_label_repel(data = filter(multipliers_adjusted, year == 2025),
                           aes(label = data_source,
                               x = 2026,
                               # fatal overdoses, MAT, nonfatal overdoses
                               y = c(9000, 35000, 1000))) +
  geom_line(aes(linetype = group)) +
  scale_linetype_manual(name = '', values = rep(c('solid', 'dashed'), each = 3)) +
  scale_color_manual(name = '', 
                     values = rep(c('firebrick3', 'orange3', 'pink3'), 2)) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
NULL
# ggsave('opioid_users_all_ADJ_multiplier_estimates.png')

### without MAT
multipliers_adjusted %>%
  filter(data_source != 'MAT') %>%

ggplot(aes(x = year, y = adj_estimate, color = group, group = group)) +
  geom_ribbon(aes(ymin = adj_low, ymax = adj_high, fill = group), 
              alpha = 0.15, color = 'white') +
  scale_fill_manual(name = '', values = rep(c('firebrick3', 'pink3'), 2)) +
  ggrepel::geom_label_repel(data = filter(multipliers_adjusted, year == 2025,
                                          data_source != 'MAT'),
                           aes(label = data_source, 
                               x = 2026, 
                               # fatal overdoses, nonfatal overdoses, MAT
                               y = c(7500, 3500))) +
  geom_line(aes(linetype = group)) +
  scale_linetype_manual(name = '', 
                        values = rep(c('solid', 'dashed'), each = 2)) +
  scale_color_manual(name = '', 
                     values = rep(c('firebrick3', 'pink3'), 2)) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
# ggsave('opioid_users_ADJ_multiplier_estimates_no_MAT.png')
```

```{r, fig.width = 5, fig.height = 4}
### all three multipliers
multipliers_adjusted %>% #distinct(group, estimate_source, data_source, adjustment)

ggplot(aes(x = year, y = estimate, color = group, group = group)) +
  geom_ribbon(aes(ymin = low, ymax = high, fill = group), 
              alpha = 0.15, color = 'white') +
  scale_fill_manual(name = '', 
                    values = rep(rep(c('firebrick3', 'orange2', 'pink3'), 
                                     each = 3), 2)) +
  # ggrepel::geom_label_repel(data = filter(multipliers_adjusted, year == 2025),
  #                          aes(label = data_source, 
  #                              x = 2026, 
  #                              # fatal overdoses, nonfatal overdoses, MAT
  #                              y = c(18000, 5000, 60000))) +
  geom_line(aes(linetype = group)) +
  scale_linetype_manual(name = '', values = rep(c('solid', 'dashed'), each = 9)) +
  scale_color_manual(name = '', 
                     values = rep(rep(c('firebrick3', 'orange2', 'pink3'), 
                                     each = 3), 2)) +
  geom_point() +
  labs(x = 'Year', y = 'Number of Individuals') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
NULL
```
